Associations between Chlorophyll *a* and various Microcystin-LR Health Advisory

### 2015-09-25 manuscript/manuscript.Rmd

In the 2007 NLA, microcystin-LR concentrations ranged from `r min(nla_dat$MCYST_TL_UGL,na.rm=T)` to `r max(nla_dat$MCYST_TL_UGL, na.rm=T)`.  Microcystin-LR concentrations of 0.05 $\mu$g/L represent the detection limits.  Any value greater than that indicates the presence of microcystin-LR.  Of those lakes with microcystin, the median concentration was `r median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T)` and the mean was `r median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T)`.  Lastly, of all lakes sampled, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=0.3])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the U.S. EPA childrens drinking water standard, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1.6])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the U.S. EPA adult drinking water standard,`r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO drinking water standard,and `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=2])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO recreational standard. For chlorophyll *a*, the range was `r min(nla_dat$CHLA,na.rm=T)` to `r max(nla_dat$CHLA, na.rm=T)`.  All lakes had reported chlorophyll *a* concentrations that exceeded detection limits.  The median concentration was `r median(nla_dat$CHLA,na.rm=T)` and the mean was `r mean(nla_dat$CHLA,na.rm=T)`.

The association between Log 10 microcystin-LR and Log 10 chlorophyll *a* show a wedge pattern (Figure \ref{fig:chla_micro_scatter}).  This indicates that higher concentrations of microcystin-LR almost always co-occur with higher concentrations of chlorophyll *a* yet the inverse is not true.  Higher chlorophyll *a* is not necessarily predictive of higher microcystin-LR concentrations; however, chlorophyll *a* may be predictive of the probability of exceeding a certian concentration.  This is the case as the probability of exceeding each of the four tested health adviroy levels shos this patern of probability of exceeding the advisory level increases as a function of chlorophyll *a* concentration (Figure \ref{fig:multi_cp_plot}).

We use this association to identify chlorophyll *a* concentrations that are associated with greater than even odds of exceeding a given health advisory level (Table \ref{tab:mc_chla_table}).  These represent `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[1]])/length(nla_dat$CHLA)),1)`%, `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[2]])/length(nla_dat$CHLA)),1)`%,
`r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[3]])/length(nla_dat$CHLA)),1)`%, and `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[4]])/length(nla_dat$CHLA)),1)`% of sample lakes for the U.S EPA childrens drinking water, the WHO drinking water, the U.S. EPA Adult drinking water, and the WHO recreational standards, respecitvely.
<!-- (Figure \ref{fig:chla_micro_cdf}) -->

#Discussion

>>>>

In the 2007 NLA, microcystin-LR concentrations ranged from `r min(nla_dat$MCYST_TL_UGL,na.rm=T)` to `r max(nla_dat$MCYST_TL_UGL, na.rm=T)`.  Microcystin-LR concentrations of 0.05 $\mu$g/L represent the detection limits.  Any value greater than that indicates the presence of microcystin-LR.  Of those lakes with microcystin, the median concentration was `r median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T)` and the mean was `r median(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>0.05],na.rm=T)`.  Lastly, of all lakes sampled, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=0.3])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the U.S. EPA childrens drinking water standard, `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1.6])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the U.S. EPA adult drinking water standard,`r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=1])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO drinking water standard,and `r round(100*(length(nla_dat$MCYST_TL_UGL[nla_dat$MCYST_TL_UGL>=2])/length(nla_dat$MCYST_TL_UGL)),1)`% of lakes exceeded the WHO recreational standard. For chlorophyll *a*, the range was `r min(nla_dat$CHLA,na.rm=T)` to `r max(nla_dat$CHLA, na.rm=T)`.  All lakes had reported chlorophyll *a* concentrations that exceeded detection limits.  The median concentration was `r median(nla_dat$CHLA,na.rm=T)` and the mean was `r mean(nla_dat$CHLA,na.rm=T)`.  The associations between chlorophyll *a* and the upper confidence interval with a conditional probability of 50% were `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]` for 0.3, 1.0, 1.6 and 2.0 $\mu$g/L, respectively (Table \ref{tab:mc_chla_table} & Figure \ref{fig:multi_cp_plot}).

#Discussion
The association between Log 10 microcystin-LR and Log 10 chlorophyll *a* show a wedge pattern (Figure \ref{fig:chla_micro_scatter}).  This indicates that higher concentrations of microcystin-LR almost always co-occur with higher concentrations of chlorophyll *a* yet the inverse is not true.  Higher chlorophyll *a* is not necessarily predictive of higher microcystin-LR concentrations; however, chlorophyll *a* may be predictive of the probability of exceeding a certain concentration.

This is the case as the probability of exceeding each of the four tested health advisory levels increases as a function of chlorophyll *a* concentration (Figure \ref{fig:multi_cp_plot}).  We use this association to identify chlorophyll *a* concentrations that are associated with greater than even odds of exceeding a given health advisory level (Table \ref{tab:mc_chla_table}).  These represent `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[1]])/length(nla_dat$CHLA)),1)`%, `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[2]])/length(nla_dat$CHLA)),1)`%,
`r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[3]])/length(nla_dat$CHLA)),1)`%, and `r round(100*(length(nla_dat$CHLA[nla_dat$CHLA>=mc_chla[4]])/length(nla_dat$CHLA)),1)`% of sample lakes for the U.S EPA childrens drinking water, the WHO drinking water, the U.S. EPA Adult drinking water, and the WHO recreational standards, respecitvely.

There are numerous possible uses for the chlorophyll *a* and microcystin-LR advisory cut-off values.  First, in the absence of microcystin-LR measurements, exceedence of the chlorophyll *a* concentrations could be a trigger for further actions.  Given that there is uncertainity around these chlorophyll *a* cutoffs the best case scenario would be to monitor for chlorophyll and in the event of exceeding a target concentration take water samples and have those samples tested in a lab for microcystin-LR.  A second potential use is to identify possible bloom events from historical data.  As harmful algal blooms are made up of many species and have various mechanism responsible for adverse impacts (e.g. toxins, hypoxia, odors), there is no single definition of a bloom.  For cHABs one approach has been to identify an increase over a baseline concentration of phycocyanin [@miller2013spatiotemporal].  This is a useful approach for targeted studies, but phycocyanin is also not always available and measures the predominance of cyanobacterial pigments and not toxins. Using our chlorophyll *a* cutoffs provides a value that is more directly associated with microcystin-LR and can be used to classify lakes, from past surveys, as having bloomed.

These values are conservative: greater than even odds for the upper CI. Most protective.

Not meant as a replacement for testing of microcystin, but provides another useful tool to help understand the probable extent of microcystin occurence.
